\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-01-07 paper.qmd
Updated Figure 11

## Compute hexbin parameters
num_bins_x_gau1 <- 20 #13
lim1 <- gau1_scaled_obj$lim1
lim2 <- gau1_scaled_obj$lim2
r2_gau1 <- diff(lim2)/diff(lim1)

@@

benchmark_gau1 <- 0.1

@@

distance_df_small_edges_gau1 <- distance_gau1 |>
  filter(distance < benchmark_gau1) |>
  filter(!(row_number() %in% c(33, 40, 48, 138, 143)))

@@

trimesh_removed_gau_tsne <- ggplot() + 
  #geom_segment(data = tr_from_to_df_gau1, 
  #             aes(
  #               x = x_from, 
  #               y = y_from, 
  #               xend = x_to, 
  #               yend = y_to),
  #             colour = "#000000",
  #             linewidth = 1) +
  geom_point(
    data = tsne_gau_scaled,
    aes(
      x = tSNE1,
      y = tSNE2
    ),
    alpha = 0.2) + 
  geom_rect(
      aes(xmin = 0.3, 
          xmax = 0.65, 
          ymin = -0.05, 
          ymax = 0.35),
          fill = "transparent", 
      color = "grey50", 
      linewidth = 0.5) +
  interior_annotation("a1", 
                      position = c(0.08, 0.95),
                      cex = 2) +
  theme(
    aspect.ratio = 1
  )

@@

axes_obj <- gen_axes(  
  proj = projection,
  limits = 0.3,
  axis_pos_x = 0.05,
  axis_pos_y = -0.85,
  axis_labels = names(scaled_gau_data),
  threshold = 0)

five_gau_proj_tsne_model1 <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#000000",
    linewidth = 1) +
  geom_point(
    #size = 0.5,
    alpha = 0.2,
    color = "#999999") +
  geom_segment(
    data=axes, 
    aes(x=x1, y=y1, xend=x2, yend=y2), 
    colour="grey70") +
  geom_text(
    data=axes, 
    aes(x=x2, y=y2),
    label=rownames(axes), 
    colour="grey50",
    size = 5) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(0, 0.4)) +
  ylim(c(-0.9, -0.5)) +
  interior_annotation("a2", 
                      position = c(0.08, 0.9),
                      cex = 2)

## Second projection
# projection <- cbind(
#     c(-0.71895,0.29250,-0.29905,-0.01656),
#     c(-0.16698,-0.62707,-0.23756,0.46328))
projection <- cbind(
  c(0.36030,0.00630,0.66545,-0.34307),
  c(-0.01861,0.14409,-0.36798,-0.73066))

axes_obj <- gen_axes(  
  proj = projection,
  limits = 0.3,
  axis_pos_x = -0.7,
  axis_pos_y = -0.7,
  axis_labels = names(scaled_gau_data),
  threshold = 0)

five_gau_proj_tsne_model2 <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#000000",
    linewidth = 1) +
  geom_point(
    #size = 0.5,
    alpha = 0.2,
    color = "#999999") +
  geom_segment(
    data=axes, 
    aes(x=x1, y=y1, xend=x2, yend=y2), 
    colour="grey70") +
  geom_text(
    data=axes, 
    aes(x=x2, y=y2), 
    label=rownames(axes), 
    colour="grey50",
    size = 5) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-0.75, -0.36)) +
  ylim(c(-0.75, -0.36)) +
  interior_annotation("a3", 
                      position = c(0.08, 0.9),
                      cex = 2)

# five_gau_proj_tsne_model <- gen_proj_langevitour(
#   points_df = df_exe,
#   projection = projection,
#   edge_df = distance_df_small_edges |> select(-distance)
# ) +
#   xlim(c(-0.18, 0.1)) +
#   ylim(c(-0.78, -0.55))
```

@@

## Compute hexbin parameters
num_bins_x_gau1 <- 45
lim1 <- gau1_scaled_obj$lim1
lim2 <- gau1_scaled_obj$lim2
r2_gau1 <- diff(lim2)/diff(lim1)

@@

distance_df_small_edges_gau1 <- distance_gau1 |>
  filter(distance < benchmark_gau1) |>
  mutate(ID = row_number()) |>
  filter(!(ID %in% c(112)))

@@

trimesh_removed_gau_umap <- ggplot() + 
  #geom_segment(data = tr_from_to_df_gau1, 
  #             aes(
  #               x = x_from, 
  #               y = y_from, 
  #               xend = x_to, 
  #               yend = y_to),
  #             colour = "#000000",
  #             linewidth = 1) +
  geom_point(
    data = umap_gau_scaled,
    aes(
      x = UMAP1,
      y = UMAP2
    ),
    alpha = 0.05) + 
  geom_rect(
    aes(xmin = 0.45, 
        xmax = 0.57, 
        ymin = 0, 
        ymax = 0.1),
    fill = "transparent", 
    color = "grey50", 
    linewidth = 0.5)  +
  interior_annotation("b1", 
                      position = c(0.06, 0.95),
                      cex = 2) +
  theme(
    aspect.ratio = 1
  )

@@

axes_obj <- gen_axes(  
  proj = projection,
  limits = 0.28,
  axis_pos_x = -0.58,
  axis_pos_y = -0.35,
  axis_labels = names(scaled_gau_data),
  threshold = 0)

five_gau_proj_umap_model1 <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#000000",
    linewidth = 1) +
  geom_point(
    #size = 0.5,
    alpha = 0.3,
    color = "#999999") +
  geom_segment(
    data=axes, 
    aes(x=x1, y=y1, xend=x2, yend=y2), 
    colour="grey70") +
  geom_text(
    data=axes, 
    aes(x=x2, y=y2, label=rownames(axes)), 
    colour="grey50",
    size = 5) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-0.64, -0.29)) +
  ylim(c(-0.4, -0.05)) +
  interior_annotation("b2", 
                      position = c(0.08, 0.9),
                      cex = 2)

## Second projection
projection <- cbind(
  c(0.36030,0.00630,0.66545,-0.34307),
  c(-0.01861,0.14409,-0.36798,-0.73066))

axes_obj <- gen_axes(  
  proj = projection,
  limits = 0.28,
  axis_pos_x = -0.7,
  axis_pos_y = -0.7,
  axis_labels = names(scaled_gau_data),
  threshold = 0)

five_gau_proj_umap_model2 <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#000000",
    linewidth = 1) +
  geom_point(
    #size = 0.5,
    alpha = 0.2,
    color = "#999999") +
  geom_segment(
    data=axes, 
    aes(x=x1, y=y1, xend=x2, yend=y2), 
    colour="grey70") +
  geom_text(
    data=axes, 
    aes(x=x2, y=y2, label=rownames(axes)), 
    colour="grey50",
    size = 5) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-0.75, -0.4)) +
  ylim(c(-0.75, -0.4)) +
  interior_annotation("b3", 
                      position = c(0.08, 0.9),
                      cex = 2)

# five_gau_proj_umap_model <- gen_proj_langevitour(
#   points_df = df_exe,
#   projection = projection,
#   edge_df = distance_df_small_edges |> select(-distance)
# ) +
#   xlim(c(-0.18, 0.1)) +
#   ylim(c(-0.78, -0.55))
```

@@

## Compute hexbin parameters
num_bins_x_gau1 <- 21
lim1 <- gau1_scaled_obj$lim1
lim2 <- gau1_scaled_obj$lim2
r2_gau1 <- diff(lim2)/diff(lim1)

@@

trimesh_removed_gau_pacmap <- ggplot() + 
  #geom_segment(data = tr_from_to_df_gau1, 
  #             aes(
  #               x = x_from, 
  #               y = y_from, 
  #               xend = x_to, 
  #               yend = y_to),
  #             colour = "#000000",
  #             linewidth = 1) +
  geom_point(
    data = pacmap_gau_scaled,
    aes(
      x = PaCMAP1,
      y = PaCMAP2
    ),
    alpha = 0.3) +
  geom_rect(
    aes(xmin = -0.04, 
        xmax = 0.22, 
        ymin = 0.08, 
        ymax = 0.34),
    fill = "transparent", 
    color = "grey50", 
    linewidth = 0.5) +
  interior_annotation("c1", 
                      position = c(0.08, 0.95),
                      cex = 2) +
  theme(
    aspect.ratio = 1
  )

@@

axes_obj <- gen_axes(  
  proj = projection,
  limits = 0.25,
  axis_pos_x = -0.41,
  axis_pos_y = 0.23,
  axis_labels = names(scaled_gau_data),
  threshold = 0)

five_gau_proj_pacmap_model1 <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#000000",
    linewidth = 1) +
  geom_point(
    #size = 0.5,
    alpha = 0.3,
    color = "#999999") +
  geom_segment(
    data=axes, 
    aes(x=x1, y=y1, xend=x2, yend=y2), 
    colour="grey70") +
  geom_text(
    data=axes, 
    aes(x=x2, y=y2, label=rownames(axes)), 
    colour="grey50",
    size = 5) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-0.46, -0.12)) +
  ylim(c(0.18, 0.52)) +
  interior_annotation("c2", 
                      position = c(0.08, 0.9),
                      cex = 2)

### Second projection
# projection <- cbind(
#     c(-0.60995,0.33361,0.32920,-0.31518),
#     c(0.19525,-0.30554,0.74417,0.07600))
projection <- cbind(
  c(0.36030,0.00630,0.66545,-0.34307),
  c(-0.01861,0.14409,-0.36798,-0.73066))

axes_obj <- gen_axes(  
  proj = projection,
  limits = 0.3,
  axis_pos_x = -0.7,
  axis_pos_y = -0.7,
  axis_labels = names(scaled_gau_data),
  threshold = 0)

five_gau_proj_pacmap_model2 <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#000000",
    linewidth = 1) +
  geom_point(
    #size = 0.8,
    alpha = 0.2,
    color = "#999999") +
  geom_segment(
    data=axes, 
    aes(x=x1, y=y1, xend=x2, yend=y2), 
    colour="grey70") +
  geom_text(
    data=axes, 
    aes(x=x2, y=y2, label=rownames(axes)), 
    colour="grey50",
    size = 5) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-0.75, -0.35)) +
  ylim(c(-0.75, -0.35))+
  interior_annotation("c3", 
                      position = c(0.08, 0.9),
                      cex = 2)

```{r}
#| label: fig-five-gau-projs
#| fig-height: 10
#| fig-width: 15
#| fig-pos: H
#| fig-cap: "NLDR's organise points in the \\gD{} layout in different ways, possibly misleadingly, illustrated using three layouts: (a) t-SNE, (b) UMAP, (c) PaCMap. The data has five Gaussian cluster in $4\\text{-}D$. The bottom row of plots shows a \\gD{} projection from a tour on $4\\text{-}D$ revealing the differences generated by the layouts on the model fits.  We would expect the model fit to be like that in (a2) where it is distinctly is separate for each cluster but like a hairball in each. This would indicate the distinct clusters, each being fully $4\\text{-}D$. With (c2), the curiousity is that the model is a \\gD{} pancake shape in $4\\text{-}D$, indicating that there is some ordering of points done by PaCMAP, posisbly along some principal component axes.  Videos of the langevitour animations are available at <https://youtu.be/oQxEb4wRdHI>, <https://youtu.be/JW49csPpDx4>, and XXX respectively."

@@

```{r}
gen_proj_five_clust()[[1]] + gen_proj_five_clust()[[2]] + plot_layout(nrow = 2)
```

>>>>

## Compute hexbin parameters
num_bins_x_gau1 <- 35
lim1 <- gau1_scaled_obj$lim1
lim2 <- gau1_scaled_obj$lim2
r2_gau1 <- diff(lim2)/diff(lim1)

@@

# benchmark_gau1 <- 0.1

@@

distance_df_small_edges_gau1 <- distance_gau1 |>
  filter(distance < benchmark_gau1)

@@

trimesh_removed_gau_tsne <- ggplot() + 
  geom_point(
    data = tsne_gau_scaled,
    aes(
      x = tSNE1,
      y = tSNE2
    ),
    alpha = 0.2) + 
  # geom_rect(
  #     aes(xmin = 0.3, 
  #         xmax = 0.65, 
  #         ymin = -0.05, 
  #         ymax = 0.35),
  #         fill = "transparent", 
  #     color = "grey50", 
  #     linewidth = 0.5) +
  interior_annotation("a1", 
                      position = c(0.08, 0.95),
                      cex = 2) +
  theme(
    aspect.ratio = 1
  )

@@

axes_obj <- gen_axes(  
  proj = projection,
  limits = 1,
  axis_pos_x = -0.6,
  axis_pos_y = -0.63,
  axis_labels = names(scaled_gau_data),
  threshold = 0)

five_gau_proj_tsne_model1 <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_point(
    #size = 0.5,
    alpha = 0.3,
    color = "#999999") +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#000000",
    alpha = 0.4,
    linewidth = 0.1) +
  geom_segment(
    data=axes, 
    aes(x=x1, y=y1, xend=x2, yend=y2), 
    colour="grey70") +
  geom_text(
    data=axes, 
    aes(x=x2, y=y2),
    label=rownames(axes), 
    colour="grey50",
    size = 5) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-0.84, 0.84)) +
  ylim(c(-0.84, 0.84)) +
  interior_annotation("a2", 
                      position = c(0.08, 0.9),
                      cex = 2)
```

@@

## Compute hexbin parameters
num_bins_x_gau1 <- 127
lim1 <- gau1_scaled_obj$lim1
lim2 <- gau1_scaled_obj$lim2
r2_gau1 <- diff(lim2)/diff(lim1)

@@

distance_df_small_edges_gau1 <- distance_gau1 |>
  filter(distance < benchmark_gau1) |>
  mutate(ID = row_number()) #|>
  #filter(!(ID %in% c(112)))

@@

trimesh_removed_gau_umap <- ggplot() + 
  geom_point(
    data = umap_gau_scaled,
    aes(
      x = UMAP1,
      y = UMAP2
    ),
    alpha = 0.05) + 
  # geom_rect(
  #   aes(xmin = 0.45, 
  #       xmax = 0.57, 
  #       ymin = 0, 
  #       ymax = 0.1),
  #   fill = "transparent", 
  #   color = "grey50", 
  #   linewidth = 0.5)  +
  interior_annotation("b1", 
                      position = c(0.06, 0.95),
                      cex = 2) +
  theme(
    aspect.ratio = 1
  )

@@

axes_obj <- gen_axes(  
  proj = projection,
  limits = 1,
  axis_pos_x = -0.6,
  axis_pos_y = -0.63,
  axis_labels = names(scaled_gau_data),
  threshold = 0)

five_gau_proj_umap_model1 <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_point(
    #size = 0.5,
    alpha = 0.3,
    color = "#999999") +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#000000",
    alpha = 0.4,
    linewidth = 0.1) +
  geom_segment(
    data=axes, 
    aes(x=x1, y=y1, xend=x2, yend=y2), 
    colour="grey70") +
  geom_text(
    data=axes, 
    aes(x=x2, y=y2, label=rownames(axes)), 
    colour="grey50",
    size = 5) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-0.83, 0.83)) +
  ylim(c(-0.83, 0.83)) +
  interior_annotation("b2", 
                      position = c(0.08, 0.9),
                      cex = 2)
```

@@

## Compute hexbin parameters
num_bins_x_gau1 <- 56
lim1 <- gau1_scaled_obj$lim1
lim2 <- gau1_scaled_obj$lim2
r2_gau1 <- diff(lim2)/diff(lim1)

@@

trimesh_removed_gau_pacmap <- ggplot() + 
  geom_point(
    data = pacmap_gau_scaled,
    aes(
      x = PaCMAP1,
      y = PaCMAP2
    ),
    alpha = 0.3) +
  # geom_rect(
  #   aes(xmin = -0.04, 
  #       xmax = 0.22, 
  #       ymin = 0.08, 
  #       ymax = 0.34),
  #   fill = "transparent", 
  #   color = "grey50", 
  #   linewidth = 0.5) +
  interior_annotation("c1", 
                      position = c(0.08, 0.95),
                      cex = 2) +
  theme(
    aspect.ratio = 1
  )

@@

axes_obj <- gen_axes(  
  proj = projection,
  limits = 1,
  axis_pos_x = -0.6,
  axis_pos_y = -0.63,
  axis_labels = names(scaled_gau_data),
  threshold = 0)

five_gau_proj_pacmap_model1 <- projected_df |>
  ggplot(
    aes(
      x = proj1,
      y = proj2)) +
  geom_point(
    #size = 0.5,
    alpha = 0.3,
    color = "#999999") +
  geom_segment(
    data = model_df,
    aes(
      x = proj1_from,
      y = proj2_from,
      xend = proj1_to,
      yend = proj2_to),
    color = "#000000",
    alpha = 0.4,
    linewidth = 0.1) +
  geom_segment(
    data=axes,
    aes(x=x1, y=y1, xend=x2, yend=y2),
    colour="grey70") +
  geom_text(
    data=axes,
    aes(x=x2, y=y2, label=rownames(axes)),
    colour="grey50",
    size = 5) +
  geom_path(
    data=circle,
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-0.83, 0.83)) +
  ylim(c(-0.83, 0.83)) +
  interior_annotation("c2",
                      position = c(0.08, 0.9),
                      cex = 2)

```{r}
#| label: fig-five-gau-projs
#| fig-height: 10
#| fig-width: 15
#| fig-pos: H
#| fig-cap: "NLDR's organise points in the \\gD{} layout in different ways, possibly misleadingly, illustrated using three layouts: (a) tSNE, (b) UMAP, (c) PaCMAP. The data has five Gaussian cluster in $4\\text{-}D$. The bottom row of plots shows a \\gD{} projection from a tour on $4\\text{-}D$ revealing the differences generated by the layouts on the model fits.  We would expect the model fit to be like that in (a2) where it is distinctly is separate for each cluster but like a hairball in each. This would indicate the distinct clusters, each being fully $4\\text{-}D$. With (c2), the curiousity is that the model is a \\gD{} pancake shape in $4\\text{-}D$, indicating that there is some ordering of points done by PaCMAP, posisbly along some principal component axes.  Videos of the langevitour animations are available at <https://youtu.be/oQxEb4wRdHI>, <https://youtu.be/JW49csPpDx4>, and XXX respectively."

@@

```{r}
# gen_proj_five_clust()[[1]] + gen_proj_five_clust()[[2]] + plot_layout(nrow = 2)
```

### 2025-01-07 paper.tex
Updated error computation of Figure 6
Updated Figure 11

\caption{\label{fig-p-d-error-in-2d-two-curvy}The \(7\text{-}D\) model
error in \gD{} layout. Color indicates square root of absolute error,
dark blue indicating high error and light indicates low error. Most
large errors are distributed near the edges of the clusters.}

@@

\caption{\label{fig-five-gau-projs}NLDR's organise points in the \gD{}
layout in different ways, possibly misleadingly, illustrated using three
layouts: (a) t-SNE, (b) UMAP, (c) PaCMap. The data has five Gaussian
cluster in \(4\text{-}D\). The bottom row of plots shows a \gD{}
projection from a tour on \(4\text{-}D\) revealing the differences
generated by the layouts on the model fits. We would expect the model
fit to be like that in (a2) where it is distinctly is separate for each
cluster but like a hairball in each. This would indicate the distinct
clusters, each being fully \(4\text{-}D\). With (c2), the curiousity is
that the model is a \gD{} pancake shape in \(4\text{-}D\), indicating
that there is some ordering of points done by PaCMAP, posisbly along
some principal component axes. Videos of the langevitour animations are
available at \url{https://youtu.be/oQxEb4wRdHI},
\url{https://youtu.be/JW49csPpDx4}, and XXX respectively.}

@@

\begin{center}
\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/unnamed-chunk-51-1.pdf}
\end{center}

>>>>

\caption{\label{fig-p-d-error-in-2d-two-curvy}The \(7\text{-}D\) model
error in \gD{} layout. Color indicates square root of total error, dark
blue indicating high error and light indicates low error. Most large
errors are distributed near the edges of the clusters.}

@@

\caption{\label{fig-five-gau-projs}NLDR's organise points in the \gD{}
layout in different ways, possibly misleadingly, illustrated using three
layouts: (a) tSNE, (b) UMAP, (c) PaCMAP. The data has five Gaussian
cluster in \(4\text{-}D\). The bottom row of plots shows a \gD{}
projection from a tour on \(4\text{-}D\) revealing the differences
generated by the layouts on the model fits. We would expect the model
fit to be like that in (a2) where it is distinctly is separate for each
cluster but like a hairball in each. This would indicate the distinct
clusters, each being fully \(4\text{-}D\). With (c2), the curiousity is
that the model is a \gD{} pancake shape in \(4\text{-}D\), indicating
that there is some ordering of points done by PaCMAP, posisbly along
some principal component axes. Videos of the langevitour animations are
available at \url{https://youtu.be/oQxEb4wRdHI},
\url{https://youtu.be/JW49csPpDx4}, and XXX respectively.}

@@
